Testing Logistic Regression Models (Supplement to: Environmental DNA Metabarcoding Reveals Winners and Losers of Global Change in Coastal Waters)

### 2020-09-09 Scripts/Supplement_ModelTesting.Rmd
adding modeltesting fit data so user can load it directly rather than rerunning; adding accompanying knitted PDF for supplement.

---
title: "Supplement: Testing Logistic Regression Models"
header-includes:
   - \usepackage{gensymb}    
   - \usepackage{fontspec}
   - \setmainfont{Open Sans}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: true
    number_sections: no
    latex_engine: xelatex
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: inline
---

```{r}
library(tidyverse)
library(rstanarm)
  options(mc.cores = parallel::detectCores())
library(loo)
library (here)
###LOAD DATA
      data.for.stan <- read_csv(here("Input","data.for.stan.csv"))

@@

@@

```{r, echo = F}
      t_prior <- student_t(df = 7, location = 0, scale = 2.5)

      #models fitting one intercept per unique area/taxon combination, plus
      #a different slope for the effect of environmental variables on each taxon.
      mod1.3vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH + Salinity | taxa) + (1 | taxonAreaIndex), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod1.2vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH | taxa) + (1 | taxonAreaIndex), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod1.1vars.pH <- stan_glmer(binary ~ 0 + (0 + pH | taxa) + (1 | taxonAreaIndex), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod1.1vars.Temperature <- stan_glmer(binary ~ 0 + (0 + Temperature | taxa) + (1 | taxonAreaIndex), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      #models fitting one intercept per taxon, plus
      #a different slope for the effect of environmental variables on each taxon. No Area term.
      mod2.3vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH + Salinity | taxa) + (1 | taxa), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod2.2vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH | taxa) + (1 | taxa), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod2.1vars.pH <- stan_glmer(binary ~ 0 + (0 + pH | taxa) + (1 | taxa), 
                              data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod2.1vars.Temperature <- stan_glmer(binary ~ 0 + (0 + Temperature | taxa) + (1 | taxa), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      #models fitting a different slope for the effect of environmental variables on each taxon, with no intercept term.
      mod3.3vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH + Salinity | taxa), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod3.2vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH | taxa), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod3.1vars.pH <- stan_glmer(binary ~ 0 + (0 + pH | taxa), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

      mod3.1vars.Temperature <- stan_glmer(binary ~ 0 + (0 + Temperature | taxa), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 2,
                             adapt_delta = 0.99,
                             refresh = 0)

@@

$$logit(p_i) = \alpha_{j} + Temperature * \beta_{1_k} + pH * \beta_{2_k}$$

@@

```{r}
stan.input <- data.for.stan %>% 
      group_by(Area) %>% 
      mutate(Temperature = vegan::decostand(Temperature, method = "standardize"),
             pH = vegan::decostand(pH, method = "standardize"),
             Salinity = vegan::decostand(Salinity, method = "standardize")) %>%
      group_by(taxa) %>%
      mutate(Noccur = sum(binary)) %>%
      filter(Noccur > 7 & Noccur < 69) %>%  #must be present or absent in at least 10% of 76 samples in dataset
      dplyr::select(-c(sample, phylum, DIC, Noccur)) %>% 
      ungroup() %>% 
      unite(Area, taxa, col = "AreaTaxa", remove = F) %>% 
      mutate(sampleIndex = 1:nrow(.),
             taxonIndex = match(taxa, unique(taxa)),
             areaIndex = match(Area, unique(Area)),
             seasonIndex = match(Season, unique(Season)),
             taxonAreaIndex = match(AreaTaxa, unique(AreaTaxa))) %>%   #unique combinations of taxon and area
      dplyr::select(-AreaTaxa)

finalModel <- stan_glmer(binary ~ 0 + (0 + Temperature + pH | taxa) + (1 | taxonAreaIndex), 
                             data = stan.input, 
                             family = "binomial",
                             prior = t_prior,
                             prior_intercept = t_prior, 
                             chains = 4,
                             adapt_delta = 0.99,
                             refresh = 0)
```

@@

```{r, echo = F}
stan.input$postpredict <- posterior_predict(finalModel) %>% colMeans()

stan.input %>% 
  left_join(data.for.stan[,c("taxa", "phylum")]) %>% 
        filter(phylum == "Bacillariophyta") %>% 
        filter(taxa %in% unique(taxa)[1:5]) %>% 
        ggplot(aes(x = pH, y = binary)) +
          geom_point() +
          geom_point(aes(x = pH, y = postpredict), color = "blue") +
          facet_grid(taxa ~ Area) +
          ylab("Prob of Presence") +
          xlab("pH (Standardized)") +
          ggtitle("Selected Diatom Taxa")

      stan.input %>% 
        filter(taxa == "Gonyaulacaceae|Alexandrium|NA") %>% 
        #filter(taxa %in% unique(taxa)[1:5]) %>% 
        dplyr::select(pH, Temperature, taxa, Area, binary, postpredict) %>% 
        pivot_longer(c(pH, Temperature)) %>% 
        ggplot(aes(x = value, y = binary)) +
        geom_point() +
        geom_point(aes(x = value, y = postpredict), color = "blue") +
        facet_grid(name ~ Area) +
        ylab("Prob of Presence") +
        xlab("pH (Standardized)") +
        ggtitle("Alexandrium")
```

>>>>

---
title: "Testing Logistic Regression Models (Supplement to: Environmental DNA Metabarcoding Reveals Winners and Losers of Global Change in Coastal Waters)"
author: "Gallego et al."
header-includes:
   - \usepackage{gensymb}    
   - \usepackage{fontspec}
   - \setmainfont{Times New Roman}
output:
  pdf_document:
    fig_caption: yes
    number_sections: no
    latex_engine: xelatex
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: inline
---

```{r, echo = F, include = F}
library(tidyverse)
library(rstanarm)
  options(mc.cores = parallel::detectCores())
library(loo)
library (here)
###LOAD DATA
      data.for.stan <- read_csv(here("Input","data.for.stan.csv"))

@@

#if loading data from previous model runs
load(file = "ModelTest_fits.Rdata")

@@

```{r, echo = F, message = F, warning = F}
#uncomment if running de novo; otherwise, load data from file, above
      # t_prior <- student_t(df = 7, location = 0, scale = 2.5)
      # 
      # #models fitting one intercept per unique area/taxon combination, plus
      # #a different slope for the effect of environmental variables on each taxon.
      # mod1.3vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH + Salinity | taxa) + (1 | taxonAreaIndex), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # mod1.2vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH | taxa) + (1 | taxonAreaIndex), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # mod1.1vars.pH <- stan_glmer(binary ~ 0 + (0 + pH | taxa) + (1 | taxonAreaIndex), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)      
      # 
      # mod1.1vars.Temperature <- stan_glmer(binary ~ 0 + (0 + Temperature | taxa) + (1 | taxonAreaIndex), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)      
      # 
      # #models fitting one intercept per taxon, plus
      # #a different slope for the effect of environmental variables on each taxon. No Area term.
      # mod2.3vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH + Salinity | taxa) + (1 | taxa), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # mod2.2vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH | taxa) + (1 | taxa), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # mod2.1vars.pH <- stan_glmer(binary ~ 0 + (0 + pH | taxa) + (1 | taxa), 
      #                         data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # mod2.1vars.Temperature <- stan_glmer(binary ~ 0 + (0 + Temperature | taxa) + (1 | taxa), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # #models fitting a different slope for the effect of environmental variables on each taxon, with no intercept term.
      # mod3.3vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH + Salinity | taxa), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # mod3.2vars <- stan_glmer(binary ~ 0 + (0 + Temperature + pH | taxa), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # mod3.1vars.pH <- stan_glmer(binary ~ 0 + (0 + pH | taxa), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)
      # 
      # mod3.1vars.Temperature <- stan_glmer(binary ~ 0 + (0 + Temperature | taxa), 
      #                        data = stan.input, 
      #                        family = "binomial",
      #                        prior = t_prior,
      #                        prior_intercept = t_prior, 
      #                        chains = 2,
      #                        adapt_delta = 0.99,
      #                        refresh = 0)      
      #

@@

$$logit(p_i) = \alpha_{j} + Temperature_i * \beta_{1_k} + pH_i * \beta_{2_k}$$

@@

```{r, echo = F, include = F}
#uncomment if running de novo; otherwise, load data from file, above
# stan.input <- data.for.stan %>% 
#       group_by(Area) %>% 
#       mutate(Temperature = vegan::decostand(Temperature, method = "standardize"),
#              pH = vegan::decostand(pH, method = "standardize"),
#              Salinity = vegan::decostand(Salinity, method = "standardize")) %>%
#       group_by(taxa) %>%
#       mutate(Noccur = sum(binary)) %>%
#       filter(Noccur > 7 & Noccur < 69) %>%  #must be present or absent in at least 10% of 76 samples in dataset
#       dplyr::select(-c(sample, phylum, DIC, Noccur)) %>% 
#       ungroup() %>% 
#       unite(Area, taxa, col = "AreaTaxa", remove = F) %>% 
#       mutate(sampleIndex = 1:nrow(.),
#              taxonIndex = match(taxa, unique(taxa)),
#              areaIndex = match(Area, unique(Area)),
#              seasonIndex = match(Season, unique(Season)),
#              taxonAreaIndex = match(AreaTaxa, unique(AreaTaxa))) %>%   #unique combinations of taxon and area
#       dplyr::select(-AreaTaxa) 
# 
# finalModel <- stan_glmer(binary ~ 0 + (0 + Temperature + pH | taxa) + (1 | taxonAreaIndex), 
#                              data = stan.input, 
#                              family = "binomial",
#                              prior = t_prior,
#                              prior_intercept = t_prior, 
#                              chains = 4,
#                              adapt_delta = 0.99,
#                              refresh = 0)
```

@@

```{r, echo = F, message = F, warning = F}
stan.input$postpredict <- posterior_predict(finalModel) %>% colMeans()

tmp <- stan.input %>% 
  left_join(data.for.stan[,c("taxa", "phylum")]) %>% 
        filter(phylum == "Bacillariophyta") %>% 
        filter(taxa %in% unique(taxa)[1:5]) 
#tmp$taxa <- gsub(".+\\|", "", tmp$taxa) #for ease of reading labels, remove family name
tmp$taxa <- gsub("\\|", "\n", tmp$taxa) #for ease of reading labels, insert newlines
tmp$taxa <- gsub(" ", "\n", tmp$taxa) #for ease of reading labels, insert newlines
        tmp %>% 
        ggplot(aes(x = pH, y = binary)) +
          geom_point() +
          geom_point(aes(x = pH, y = postpredict), color = "blue") +
          geom_smooth(aes(x = pH, y = postpredict), color = "blue", method = "lm") +
          facet_grid(taxa ~ Area) +
          theme(aspect.ratio=1/3) +
          ylab("Prob of Presence") +
          xlab("pH (Standardized)") +
          ggtitle("Selected Diatom Taxa")

```{r, echo = F}
      stan.input %>% 
        filter(taxa == "Gonyaulacaceae|Alexandrium|NA") %>% 
        #filter(taxa %in% unique(taxa)[1:5]) %>% 
        dplyr::select(pH, Temperature, taxa, Area, binary, postpredict) %>% 
        pivot_longer(c(pH, Temperature)) %>% 
        ggplot(aes(x = value, y = binary)) +
        geom_point() +
        geom_point(aes(x = value, y = postpredict), color = "blue") +
        geom_smooth(aes(x = value, y = postpredict), color = "blue", method = "lm") +
        facet_grid(name ~ Area) +
        ylab("Prob of Presence") +
        xlab("pH (Standardized)") +
        ggtitle("Alexandrium")
```

```{r Store Data if Desired, echo  = F, include = F}
#save.image(file = "ModelTest_fits.Rdata")
```
